Keep your Laziness in Check

### 2018-01-09 paper/main.tex
More work on the intro.

\begin{itemize}
\item Laziness is a fundamental feature in Haskell \leo{WhyFPmatters?}
\item However, reasoning about programs with laziness (performance, behavior) is hard.
\item It is very easy to use a too-lazy or too-strict implementation that yields different results.
\item FRP quote here?
\item
\item Example: folds.
\item \lk{fold} to calculate sum breaks for large lists because of thunks
\item \lk{fold'} is good for some, but not for average because of whnf.
\item We need an implementation that is strict in the pair as well.
\item
\item Which leads to the natural question, how can we test the laziness behavior of functions?
\item Even worse, how can we even {\em specify} the laziness behavior?
\end{itemize}

Our technical contributions are as follows:
\begin{itemize}
\item We introduce a novel approach to specifying the laziness behavior of functions that is
{\em precise} \leo{define}. \leo{Also add drawbacks/partiality here? If not here, where in the intro?}
\item We propose a random testing approach to test such specifications.
\item We implement StrictCheck. \leo{In order for these last two points to be different, we need
something about the cool/awesome hacks in the implementation to point out here}
\item As a side contribution? We fix the function generaton of QuickCheck to produce lazy/partial functions instead
of fully strict ones all the time.
\end{itemize}
%
Section~\ref{sec:related} discusses related work. We conclude and draw
directions for future work in Section~\ref{sec:concl}.

>>>>

\leo{I want to keep a counter on how many times we'll rewrite this intro :)}
\leo{We *have* to mention Haskell in this list. Maybe with John's Why FP matters?}
Laziness is ubiquitous in modern software engineering: the big-data
framework Spark uses laziness to drive computations on demand
(\hzh{find references to spark}), Apple's new programming language
Swift includes lazy evaluation for initializing expensive resources as
needed\cn, and many functional reactive programming web frameworks rely
on it\cn. 
%
Unfortunately, reasoning about the performance of lazy programs with
laziness is hard. It is very easy\leo{I don't like 'easy' here. It has
a slightly bad antithesis with the previous 'hard' as it doesn't refer
to the same thing} to accidentally make a program too lazy or too
strict, which in turn makes the program behave differently than 
the programmer had intended.

\leo{Didn't we have an FRP quote by someone that Laziness was the third source of
bugs or something? Work that in?}

\hzh{Since laziness is a fundamental feature of the programming language Haskell, we
will explore how to observe and specify laziness using just the features
provided by Haskell, without custom compiler support.}
\leo{It feels more natural to go directly into the example that explains this difficulty.
Mentioning haskell needs to happen earlier, maybe in the original
list. Or in a ``starting with Haskell'' way}

For example, consider the following simple Haskell implementation of the \lk{sum} function:
\leo{I created an environment for this so that we can tweak them all as needed}
\begin{inlinecode}
sum = foldl (+) 0
\end{inlinecode}
%
Although this seemingly innocent implementation of sum will produce correct
results on small lists of numbers, it will quickly run out of memory on larger
lists. The culprit here is the strictness behavior of \lk{foldl}---it does not
force the accumulated result during the fold. Instead, calling this \lk{sum} on a list
produces a thunk that is linear in the size of of the input list.
%
We can fix this space leak for \lk{sum} using a strict fold, \lk{foldl'},
which forces the accumulator.

However, even this is not enough in general. Consider a
single-traversal computation of an average where we keep both a
running sum and a running length in an accumulator.
%
\begin{inlinecode}
avg = uncurry (/) . foldl (\(s,n) x -> (s + x, n + 1)) (0,0)
\end{inlinecode}
%
In this case, both \lk{foldl} and \lk{foldl'} construct thunks linear in the
size of the input list, since forcing a pair into weak head normal form does not
force either component\leo{of the pair?}.

How, then, can we test and catch these laziness bugs before they
manifest themselves in production? Is it possible to precisely specify
laziness behavior just as we would functional correctness?
%
\leo{Rewrite the next.}
In this paper, we demonstrate how to observe laziness from within
Haskell itself, and how to write down a specification for laziness
also using Haskell.

Our technical contributions are as follows:
\begin{itemize}
\item We introduce a novel approach to specifying the laziness behavior of functions that is
{\em precise} \leo{define}. \leo{Also add drawbacks/partiality here? If not here, where in the intro?}
\item We propose a random testing approach to test such specifications.
\item We demonstrate non-trivial generic programming techniques to apply StrictCheck to almost all data types. \leo{In order for these last two points to be different, we need
something about the cool/awesome hacks in the implementation to point out here}
\item As a side contribution? We fix the function generaton of QuickCheck to produce lazy/partial functions instead
of fully strict ones all the time.
\end{itemize}
%
Section~\ref{sec:related} discusses related work. We conclude and draw
directions for future work in Section~\ref{sec:concl}.
